Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-21 demystify.tex
Updates from Overleaf
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might also be useful for \emph{domestic students}.
Moreover, while this handbook is primarily intended for students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

In addition, this handbook can help \textbf{US faculty and admission committee} gain a better understanding of international students and their cultural differences.  By recognizing and leveraging these differences, CS programs in the US can attract larger and more competitive application pools from international students.

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

The guideline can be useful for all students, but it is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information presented here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I also hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

While this document may not fully meet the needs of students aiming for top-10 schools, which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other R1 reputable institutions. It is certainly useful for students aiming for GMU!

@@

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by faculty from the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice does that for PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

% \begin{commentbox}[Hung]
%   The application fee varies between schools, and is about \$70--\$100. I found this number pretty ridiculous. In most cases, there is no proper justification for that amount. Yes, the application fee could act as a filter, but a nominal amount would do the job. While a blanket waiver might not be feasible, universities should extend the eligibility of fee waiver applications to more countries.
% \end{commentbox}

@@

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should have detail, enthusiasm, and specific examples of the applicant's achievements, all of which paint a compelling picture of the candidate to the adcom.

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often send them an email or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to them. I don't think this bothers them a bit; I think they are actually interested in these ``dramas'' in my life and would be offended if I didn't come to them.

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might also be useful for \emph{US domestic students}.
Moreover, while this is primarily intended for students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

This information can also help \textbf{US faculty and admission committee} gain a better understanding of international students and their cultural differences.  By recognizing and leveraging these differences, CS programs in the US can attract larger and more competitive application pools from international students.

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

The guideline can be useful for all students, but it is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from US universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I also hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

While this document may not fully meet the needs of students aiming for top-10 schools, which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for GMU!

@@

After you submit your PhD application (usually around December of year $X-1$ for the Fall admission of year $X$), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

@@

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should give details of the applicant's research background and achievements, all of which paint a compelling picture of the candidate to the adcom.

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often send them an email or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to them. I think this does not bother them a bit and that they are actually interested in these ``dramas'' in my life.
